\mutagen: Coverage-Guided, Property-Based Testing using Exhaustive Structure-Preserving Mutations

### 2022-03-02 papers/fse22/main.tex
Compressing Background

In this work, we establish several aspects of the original CGPT approach by
\citeauthor{lampropoulos2019coverage} that leave considerable room for
improvement (see \S \ref{sec:background}).
%
%
In particular:
%
\begin{inparaenum}
\item automatically derived structure-preserving mutators can turn out
"shallow", ineffective at transforming deep test cases more than superficially;
  % superficial and rely heavily on randomness to produce diverse mutants,
\item the queuing mechanism can cause delays if interesting test cases are
  enqueued frequently and there is no way to prioritize them; and
\item the heuristic used to assign a ``mutation budget'' to each interesting
  test case (often referred to as a \emph{power schedule}) requires fine tuning
  and can be hard to generalize.
%to different kinds of testing properties.
\end{inparaenum}
%
\footnote{Some of these limitations were already identified by the authors in
their original work.}
%
%% Moreover, we replicated the Information-Flow Control (IFC) stack machine case
%% study bundled with \fuzzchick and observed a surprising limitation:
%% %
%% after repeating each experiment 30 times, \fuzzchick could only find 5 out of
%% the 20 injected bugs with a 100\% efficacy.
%% %
%% %% the hardest one being found only around 13\% of the time after an hour of
%% %% testing.
%% %
%% Despite this, \citeauthor{lampropoulos2019coverage} show that \fuzzchick is far
%% better than using normal random testing coupled with na\"ive random generators
%% by comparing the mean-time-to-failure (MTTF) against \emph{QuickChick}.
%% %
%% However, we believe that a more meticulous evaluation ought to take failure to
%% find a counterexample as an important metric when comparing PBT tools, not just
%% mean time to failure --- when one is found.

@@

%% Validation
We validate our ideas using two case studies (described in detail in \S
\ref{sec:casestudies}): the Information-Flow Control (IFC) stack machine used by
\citeauthor{lampropoulos2019coverage} in their original work, and an existing
{Web-Assembly} engine of industrial strength written in Haskell.
%
In both case studies, we compare the effect of the heuristics implemented on top
of the base testing loop of \mutagen.
%
In the IFC stack machine case study, we compare \mutagen against \fuzzchick, the
reference CGPT implementation developed in Coq \cite{lampropoulos2019coverage}.
%
Our results (\S \ref{sec:evaluation}) indicate that \mutagen outperforms
\fuzzchick both in terms of time to failure and failure rate.
%
On the other hand, our WebAssembly case study shows the performance of our tool
in a realistic scenario.
%
There, \mutagen is capable of reliably finding 15 planted bugs in the validator
and interpreter, as well as 3 previously unknown ones.
% bugs that flew under the radar
%of the existing unit test suite of this engine.
%
% This case study also help us to compare the performance of our tool against a
% more traditional RPBT approach that does not rely on code instrumentation.
In this case study we also compare the performance versus the overhead of our
tool (and our custom code instrumentation mechanism) by comparing it against
\quickcheck (black-box).  
%
All in all, our evaluation encompasses more than 600 hours of computing time and
indicates that \emph{testing mutants exhaustively together with our heuristics
to escape scalability issues} can be an appealing technique for finding bugs
reliably without sacrificing speed.%
%
\footnote{A replication package \cite{anon_2021} is available for reviewing
  purposes.}

@@

In this section, we briefly describe the motivation, ideas and limitations
behind CGPT \cite{lampropoulos2019coverage}.
%
To illustrate this technique, let us focus on a simple property defined over
binary trees.
%
Such a data structure can be defined in Haskell using a custom algebraic data
type with two data constructors for leaves and branches respectively:

@@

\noindent The type parameter \texttt{a} indicates that trees can be instantiated
using any other type as payload, so the value \texttt{Leaf True} has type
\texttt{Tree Bool}, whereas the value \texttt{Branch (Leaf 1) 2 (Leaf 3)} has
type \texttt{Tree Int}.
%
%% Then, we can define tree reflections using a simple recursive function that
%% pattern matches against the two possible constructors, inverting the order of
%% the subtrees whenever it encounters a branch:
%
%% \begin{verbatim}
%% mirror :: Tree a -> Tree a
%% mirror (Leaf x)       = Leaf x
%% mirror (Branch l x r) = Branch (mirror r) x (mirror l)
%% \end{verbatim}
%
%% Later, a reasonable requirement to assert for is that \texttt{mirror} must be
%% \emph{involutive}, i.e., reflecting a tree twice always yields the original
%% tree.
%% %
%
Then, if we assume existence of a function \texttt{balanced} of type
\texttt{Tree a -> Bool} that asserts that a tree satisfies a certain notion of
balancedness, we can write testing properties to assert that the operations
defined over the \texttt{Tree} data type preserve this invariant.
%
For instance, the following testing property can be used to find bugs in the
implementation of a given \texttt{insert} function:

@@

\noindent This property asserts that, given an element \texttt{x} and a balanced
tree \texttt{t} as input, inserting \texttt{x} into \texttt{t} will produce a
balanced tree as output.
%
(The implementation details of \texttt{balanced} and \texttt{insert} are not
important here.)
%
Moreover, the arrow operator (\texttt{==>}) indicates that \texttt{balanced t}
is a precondition of this property, and so attempting to test it using an
unbalanced tree as input will result in the test case getting \emph{discarded}.

The last missing piece is a random generator of trees.
%
%% In \quickcheck, this is commonly done via the type class mechanism
%% \cite{jones1997type}, instantiating the \texttt{Arbitrary} type class with a
%% random generator as the implementation of the overloaded \texttt{arbitrary}
%% operation:
We can write a na\"ive generator for trees of integers (simplified here to make it
more accessible) as follows:
%% \footnote{In practice, this code will most likely be implemented using Haskell's
%%   type-class mechanism, hiding the complexity behind the size limit and the
%%   payload generator.}%

@@

\noindent This definition follows a common type-directed generation approach,
and it is a good example of what to expect from a random generator synthesized
automatically from the codebase.
%
In esence, this generator randomly picks one data constructor from its
corresponding data type definition with uniform probability (i.e, either a
\texttt{Leaf} or a \texttt{Branch}), and then calls itself for every recursive
subexpression needed to produce a well-typed value, carefully reducing the input
size limit \texttt{size} by a unit on each step, ensuring termination by
restricting itself to generating leaves when it reaches zero (case \texttt{size
  == 0}).
%
Integers are generated simply by delegating the task to an external random
generator (\texttt{genInt}) defined elsewhere.
%

@@

Readers familiar with RBPT will recognize that using \texttt{genTree} to test
\texttt{prop\_insert\_inv} with \quickcheck directly does not work well.
%
In practice, \quickcheck gives up if most of the generated test cases get
discarded due to failed preconditions.
%
%because they do not satisfy the balanced precondition.
%
Sadly, most of the inputs produced by our na\"ive generator suffer from this
problem, and the interesting part of the property (\texttt{balanced (insert x
  t)}) is tested very sporadically as a result.
%% %
%% In other words, \emph{the precondition of our testing property becomes very
%%   sparse when we use simple, automatically derived generators.}

The next subsection shows how GCPT helps overcoming this limitation without
having to write specialized generators by hand.

@@

Using code instrumentation in tandem with mutations is a well-known technique in
the fuzzing community.
%
Generic fuzzing tools like \emph{AFL} \cite{afl}, \emph{HonggFuzz}
\cite{honggfuzz}, or \emph{libFuzzer} \citeyearpar{libfuzzer} as well as
language-specific ones like Crowbar \cite{dolan2017testing} or \emph{Kelinci}
\cite{kersten2017poster} use execution traces to recognize interesting test
cases, e.g, those that exercise previously undiscovered parts of the target
code.
%
Later, such tools use generic mutators to combine and produce new test cases
from previously executed interesting ones.
%
%% Instead of mutating any previously executed test case that discovers a new part
%% of the code, this technique integrates these fuzzing mechanisms into the PBT
%% testing loop itself.
%
In contrast with traditional fuzzing approaches, CGPT can distinguish
semantically valid test cases from invalid ones, i.e., those passing the testing
property preconditions as opposed to those that are discarded early.
%
This information is used to favour mutating valid test cases over discarded
ones.

>>>>

In this work, we establish several aspects%
%
\footnote{Some of these limitations were already identified by the authors in
their original work.}
%
of the original CGPT approach by \citeauthor{lampropoulos2019coverage} that
leave considerable room for improvement (see \S \ref{sec:background}).
%
In particular:
%
\begin{inparaenum}
\item automatically derived structure-preserving mutators can turn out
"shallow", ineffective at transforming deep test cases more than superficially;
  % superficial and rely heavily on randomness to produce diverse mutants,
\item the queuing mechanism can cause delays if interesting test cases are
  enqueued frequently and there is no way to prioritize them; and
\item the heuristic used to assign a ``mutation budget'' to each interesting
  test case (often referred to as a \emph{power schedule}) requires fine tuning
  and can be hard to generalize.
%to different kinds of testing properties.
\end{inparaenum}
%
%% Moreover, we replicated the Information-Flow Control (IFC) stack machine case
%% study bundled with \fuzzchick and observed a surprising limitation:
%% %
%% after repeating each experiment 30 times, \fuzzchick could only find 5 out of
%% the 20 injected bugs with a 100\% efficacy.
%% %
%% %% the hardest one being found only around 13\% of the time after an hour of
%% %% testing.
%% %
%% Despite this, \citeauthor{lampropoulos2019coverage} show that \fuzzchick is far
%% better than using normal random testing coupled with na\"ive random generators
%% by comparing the mean-time-to-failure (MTTF) against \emph{QuickChick}.
%% %
%% However, we believe that a more meticulous evaluation ought to take failure to
%% find a counterexample as an important metric when comparing PBT tools, not just
%% mean time to failure --- when one is found.

@@

%% Validation
We validated our ideas using two case studies (described in detail in \S
\ref{sec:casestudies}): the Information-Flow Control (IFC) stack machine used by
\citeauthor{lampropoulos2019coverage} in their original work, and an existing
{Web-Assembly} engine of industrial strength written in Haskell.
%
In both case studies, we compare the effect of the heuristics implemented on top
of the base testing loop of \mutagen.
%
In the IFC stack machine case study, we compare \mutagen against \fuzzchick, the
reference CGPT implementation developed in Coq \cite{lampropoulos2019coverage}.
%
Our results (\S \ref{sec:evaluation}) indicate that \mutagen outperforms
\fuzzchick both in terms of time to failure and failure rate.
%
On the other hand, our WebAssembly case study shows the performance of our tool
in a realistic scenario.
%
There, \mutagen is capable of reliably finding 15 planted bugs in the validator
and interpreter, as well as 3 previously unknown ones.
% bugs that flew under the radar
%of the existing unit test suite of this engine.
%
% This case study also help us to compare the performance of our tool against a
% more traditional RPBT approach that does not rely on code instrumentation.
In this case study we also compare the performance versus the overhead of our
tool (and our custom code instrumentation mechanism) by comparing it against
\quickcheck (black-box).  
%
All in all, our evaluation encompasses more than 600 hours of computing time and
indicates that \emph{testing mutants exhaustively together with our heuristics
to escape scalability issues} can be an appealing technique for finding bugs
reliably without sacrificing speed.%
%
\footnote{A replication package \cite{anon_2021} is available for reviewing
  purposes.}

@@

This section briefly introduces the motivation, ideas and limitations behind
CGPT \cite{lampropoulos2019coverage}.
%
To illustrate this technique, we focus on a simple property defined over binary
trees.
%
Such a data structure can be defined in Haskell using a custom algebraic data
type with two data constructors for leaves and branches respectively:

@@

\noindent The type parameter \texttt{a} indicates that trees can be instantiated
using any type as payload, so the value \texttt{Leaf True} has type \texttt{Tree
Bool}, whereas the value \texttt{Branch (Leaf 1) 2 (Leaf 3)} has type
\texttt{Tree Int}.
%
%% Then, we can define tree reflections using a simple recursive function that
%% pattern matches against the two possible constructors, inverting the order of
%% the subtrees whenever it encounters a branch:
%
%% \begin{verbatim}
%% mirror :: Tree a -> Tree a
%% mirror (Leaf x)       = Leaf x
%% mirror (Branch l x r) = Branch (mirror r) x (mirror l)
%% \end{verbatim}
%
%% Later, a reasonable requirement to assert for is that \texttt{mirror} must be
%% \emph{involutive}, i.e., reflecting a tree twice always yields the original
%% tree.
%% %
%
If we assume existence of a function \texttt{balanced} of type \texttt{Tree a ->
Bool} that asserts that a tree satisfies a certain notion of balancedness, we
can write testing properties to validate that the operations defined over the
\texttt{Tree} data type preserve this invariant.
%
For instance, the following can be used to validate the implementation of an
\texttt{insert} function:

@@

\noindent This property asserts that, given an element \texttt{x} and a balanced
tree \texttt{t} as input, inserting \texttt{x} into \texttt{t} will produce a
balanced tree as output.
%
(The implementation details of \texttt{balanced} and \texttt{insert} are not
important here.)
%
The arrow operator (\texttt{==>}) indicates that \texttt{balanced t} is a
precondition of this property, and so using an unbalanced tree as input will
result in the test case getting \emph{discarded}.

The only missing piece is a random generator of trees.
%
For this, we can define a na\"ive generator for trees of integers as:
%
%% In \quickcheck, this is commonly done via the type class mechanism
%% \cite{jones1997type}, instantiating the \texttt{Arbitrary} type class with a
%% random generator as the implementation of the overloaded \texttt{arbitrary}
%% operation:
%% \footnote{In practice, this code will most likely be implemented using Haskell's
%%   type-class mechanism, hiding the complexity behind the size limit and the
%%   payload generator.}%

@@

\noindent This definition (simplified to make it more accessible) follows a
common type-directed approach, and illustrates what to expect from a generator
synthesized automatically from the codebase.
%
At each step, this generator randomly picks a \texttt{Tree} data constructor
with uniform probability (i.e, either a \texttt{Leaf} or a \texttt{Branch}), and
calls itself for every recursive subexpression, carefully reducing the input
size limit \texttt{size} by a unit at a time.
%
This enforces termination by generating only leaves when the size reaches zero
(case \texttt{size == 0}).
%
Integers are generated simply by delegating the task to an external random
generator (\texttt{genInt}) defined elsewhere.
%

@@

Readers familiar with RBPT will recognize that using \texttt{genTree} to test
\texttt{prop\_insert\_inv} with \quickcheck directly does not work.
%
In practice, \quickcheck gives up if most of the generated test cases get
discarded.
%
%because they do not satisfy the balanced precondition.
%
Sadly, most of the test cases produced by \texttt{genTree} consist of unbalanced
trees, thus the interesting part of the property (\texttt{balanced (insert x
t)}) is rarely exercised.
%% %
%% In other words, \emph{the precondition of our testing property becomes very
%%   sparse when we use simple, automatically derived generators.}

% The next subsection shows how GCPT helps overcoming this limitation without
% having to write specialized generators by hand.

@@

Using code instrumentation in tandem with mutations is a well-known technique in
the fuzzing community.
%
Generic fuzzing tools like \emph{AFL} \cite{afl}, \emph{HonggFuzz}
\cite{honggfuzz}, or \emph{libFuzzer} \citeyearpar{libfuzzer} as well as
language-specific ones like Crowbar \cite{dolan2017testing} or \emph{Kelinci}
\cite{kersten2017poster} use execution traces to recognize interesting test
cases, 
%
% e.g, those that exercise previously undiscovered parts of the target
% code.
%
Later, such tools use generic mutators to combine and produce new test cases
from previously executed interesting ones.
%
%% Instead of mutating any previously executed test case that discovers a new part
%% of the code, this technique integrates these fuzzing mechanisms into the PBT
%% testing loop itself.
%
In contrast with traditional fuzzing approaches, CGPT can distinguish
semantically valid test cases from invalid ones, i.e., those passing the testing
property preconditions as opposed to those that are discarded early.
%
This information is used to favour mutating valid test cases over discarded
ones.
